---
title: "Thesis R code"
output:
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(getwd())
knitr::opts_chunk$set(cache.rebuild = TRUE)
# handle some engines with their own custom routines 
```

## Preface

This markdown document contains the code used for the thesis entitled "A Double Machine Learning approach to estimate the causal effects of behavior, beliefs and attitudes about Covid-19 on antibody status".
Author: Charley Bosman

## Import data and load packages
```{r loading data and packages, include=FALSE}
library(causalweight)
library(dplyr)
library(tidyr)
library(psych)
library(mediation)
library(plyr)
library(ggplot2)
library(naniar)
library(psych)
library(performanceEstimation)
library(curl)
library(googledrive)
x <- drive_get("data.csv") # get ID
id <- "1vTqkm58NYhjp-VBWKMnBdnNu7THhXo0f" # google file ID
data <- read.csv(sprintf("https://docs.google.com/uc?id=%s&export=download", id), stringsAsFactors=TRUE)
```

```{r self made functions}
my_summary <- function(v){
  if(!any(is.na(v))){
    res <- c(summary(v),"NA's"=0)
  } else{
    res <- summary(v)
  }
  return(res)
}

reverse <- function(x){
  x = 6-x
  return(x)
}

reverse1 <- function(x){
  x = 1-x
  return(x)
}

normalize <- function(x) { 
  x <- as.matrix(x)
  minAttr=apply(x, 2, min)
  maxAttr=apply(x, 2, max)
  x <- sweep(x, 2, minAttr, FUN="-") 
  x=sweep(x, 2,  maxAttr-minAttr, "/") 
  attr(x, 'normalized:min') = minAttr
  attr(x, 'normalized:max') = maxAttr
  return (x)
} 
```

## Cleaning
```{r drop NAs of covid result}
data <- data %>%
  filter(Covid_result=="neg"|Covid_result=="pos") %>%
  droplevels()
table(data$Covid_result)
newData <- data
```
```{r cleaning, removing outliers, detect missing vlalues}
newData <- newData %>% select(-c(Extra_Anders_text, Extra_Niets))

newData <- newData %>% mutate_at(vars(starts_with("Klacht")), ~replace(., is.na(.), 0))
newData <- newData %>% mutate_at(vars(starts_with("Extra")), ~replace(., is.na(.), 0))

newData <- newData %>% mutate(sumExtra = select(., starts_with("Extra")) %>% rowSums())

newData <- newData %>% mutate(sumKlacht = select(., starts_with("Klacht")) %>% rowSums())

gg_miss_var(newData)
gg_miss_upset(newData)
gg_miss_upset(newData, nsets = n_var_miss(newData))

newData <- newData %>% select(-c(Covid_result_repeat, ResponseType, Huishouden_anders, ThuiswonendeKinderen_0.3, ThuiswonendeKinderen_4.12, ThuiswonendeKinderen_13.17, ThuiswonendeKinderen_18., DirectContact, DirectContactMetCovid,HoeveelPersonenContact, HoeContact, DatumGetest, TweedeOnderzoek, contains("Donor"), contains("Doner"), StartDate, EndDate, BeginKlachten, EindKlachten, ArtsGeweest, Getest))

gg_miss_var(newData)

boxplot(newData$MinutenAandacht)
outliers <- boxplot(newData$MinutenAandacht,plot=FALSE)$out
newData <- newData[-which(newData$MinutenAandacht %in% outliers),]
boxplot(newData$MinutenAandacht)

boxplot(newData$TotaleTijd)
outliers <- boxplot(newData$TotaleTijd,plot=FALSE)$out
newData <- newData[-which(newData$TotaleTijd %in% outliers),]
boxplot(newData$TotaleTijd)
```

```{r smote method and delete NAs}
newData <- newData[complete.cases(newData), ]
newData <- smote(Covid_result ~ ., newData, perc.over = 5.5, perc.under = 1.2)
table(newData$Covid_result)
```

```{r rescaling stellingen}
hist(newData$VertrouwenArtsen)
newData$VertrouwenArtsen <- reverse(newData$VertrouwenArtsen)
hist(newData$VertrouwenArtsen)
hist(newData$EerlijkOverheid)
newData$EerlijkOverheid <- reverse(newData$EerlijkOverheid)
hist(newData$EerlijkOverheid)
hist(newData$VertrouwenGezondheidssysteem)
newData$VertrouwenGezondheidssysteem <- reverse(newData$VertrouwenGezondheidssysteem)
hist(newData$VertrouwenGezondheidssysteem)
hist(newData$VertrouwenOverheid)
newData$VertrouwenOverheid <- reverse(newData$VertrouwenOverheid)
hist(newData$VertrouwenOverheid)
```

## Creating new columns to summarize values
```{r creating new columns}
newData <- newData %>% mutate(gedragRIVM = rowMeans(select(newData, starts_with("RIVM")), na.rm = TRUE))

newData <- newData %>% mutate(gedragExtra = select(., starts_with("Extra")) %>% rowSums())

newData <- newData %>% mutate(effectiefmaatregelen = rowMeans(select(newData, starts_with("Effectief")), na.rm = TRUE))

newData <- newData %>% mutate(bezorgd = rowMeans(select(newData, c(starts_with("Bezorgd"), Stelling_SerieuzeZiekte, Stelling_ProblemenOmgeving, Stelling_GroteGevolgen, Stelling_Gespannen, Stelling_Zorgen, Stelling_ZorgenFamilie)), na.rm = TRUE))

newData <- newData %>% mutate(vertrouwen = rowMeans(select(newData, c("VertrouwenOverheid", "EerlijkOverheid", "VertrouwenGezondheidssysteem", "VertrouwenArtsen")), na.rm = TRUE))

means <- newData %>% group_by(Covid_result) %>% summarise_at(c("bezorgd", "gedragRIVM", "gedragExtra", "vertrouwen", "effectiefmaatregelen"), mean, na.rm=TRUE)

control <- newData %>% select(-c(Covid_result, bezorgd, gedragRIVM, gedragExtra, sumExtra, effectiefmaatregelen, vertrouwen, ID, VertrouwenArtsen, VertrouwenGezondheidssysteem, VertrouwenOverheid, EerlijkOverheid, starts_with("Bezorgd"), starts_with("Effectief"), starts_with("Extra"), starts_with("RIVM")))

summary <- newData %>% select(c(Geslacht, Leeftijd, HoogsteOpleiding, sumKlacht, Covid_result, bezorgd, gedragRIVM, gedragExtra, effectiefmaatregelen, vertrouwen, OD.CO, CoronaWeek_Thuiswerken, CoronaWeek_Buitenshuiswerk, CoronaWeek_Nietwerken, NormaleWeek_Thuiswerken, NormaleWeek_Buitenshuiswerken, Normaleweek_Nietwerken, CruciaalBeroep, MinutenAandacht, Stelling_OverspoeldMetInformatie)) %>% drop_na()

summary$Geslacht <- revalue(summary$Geslacht, c("F"="Female", "M"="Male"))
```

```{r normalize and rescale}
sum_norm <- summary %>% mutate_at(vars(c(bezorgd, gedragRIVM, gedragExtra, effectiefmaatregelen, vertrouwen)), funs(normalize))

sum_norm <- sum_norm %>% mutate_at(vars(c(bezorgd, gedragRIVM, , effectiefmaatregelen, vertrouwen)), funs(reverse1))

hist(sum_norm$gedragExtra)
hist(sum_norm$gedragRIVM)
hist(summary$gedragExtra)
sum_norm <- sum_norm %>% mutate(gedrag = (gedragRIVM+gedragExtra)/2)
hist(sum_norm$gedrag)
```
## Exploratory data analysis

```{r  exploratory analysis}
sum_norm <- sum_norm %>% mutate(Leeftijd_cat = cut(Leeftijd, breaks=c(-Inf, 30, 50, Inf), labels=c("<30","30-50",">50")))

meanscovid <- sum_norm %>% group_by(Covid_result) %>% summarise_at(c("bezorgd", "gedragRIVM", "gedragExtra", "gedrag", "vertrouwen", "effectiefmaatregelen"), mean, na.rm=TRUE)

meansage <- sum_norm %>% group_by(Leeftijd_cat, Covid_result, Geslacht) %>% summarise_at(c("bezorgd", "gedragRIVM", "gedragExtra", "gedrag", "vertrouwen", "effectiefmaatregelen"), mean, na.rm=TRUE)

meansgender <- sum_norm %>% group_by(Geslacht, Covid_result) %>% summarise_at(c("bezorgd", "gedragRIVM", "gedragExtra", "gedrag", "vertrouwen", "effectiefmaatregelen"), mean, na.rm=TRUE)

meansall <- newData %>% group_by(Covid_result) %>% summarise_if(is.numeric, mean)
summary(sum_norm)
summary(summary)
summary(meansgender)
summary(meansage)
summary(meanscovid)
summary(newData$Leeftijd)
```

## Plotting
```{r plotting histograms}
mu <- ddply(sum_norm, "Covid_result", summarise, meangedrag=mean(gedrag), meanbezorgd=mean(bezorgd), meanvertrouwen=mean(vertrouwen), effectiefmaatregelenmean=mean(effectiefmaatregelen))
head(mu)

freqbehavior <- ggplot(sum_norm, aes(x = gedrag, fill=Covid_result)) + geom_density(alpha=0.7) + 
  geom_vline(data=mu, aes(xintercept = meangedrag, color=Covid_result), linetype="solid", size=1.5) +
  scale_x_continuous(labels = c("Not so strict", "", "", "", "Very strict"), name = "Behavior") +
  scale_y_continuous(name="Frequency", limits = c(0,4)) +  scale_fill_brewer(palette="Set1") +
  scale_color_brewer(palette="Set1") + theme_classic() + labs(fill = "Covid result", color = "Covid result") +
  theme(axis.ticks.y = element_blank(), axis.text.y=element_blank(), legend.position = "top", plot.margin =
          unit(c(1, 1, 0, 0), "cm"))

freqtrust <- ggplot(sum_norm, aes(x = vertrouwen, fill=Covid_result)) + geom_density(alpha=0.7) + 
  geom_vline(data=mu, aes(xintercept = meanvertrouwen, color=Covid_result), linetype="solid", size=1.5) +
  scale_x_continuous(labels = c("No trust", "", "", "", "A lot of trust"), name = "Trust") +
  scale_y_continuous(name="Frequency", limits = c(0,4)) +  scale_fill_brewer(palette="Set1") +
  scale_color_brewer(palette="Set1") + theme_classic() + labs(fill = "Covid result", color = "Covid result") +
  theme(axis.ticks.y = element_blank(), axis.text.y=element_blank(), legend.position = "top", plot.margin =
          unit(c(1, 1, 0, 0), "cm"))

freqeffective <- ggplot(sum_norm, aes(x = effectiefmaatregelen, fill=Covid_result)) + geom_density(alpha=0.7) + 
  geom_vline(data=mu, aes(xintercept = effectiefmaatregelenmean, color=Covid_result), linetype="solid", size=1.5) +
  scale_x_continuous(labels = c("Not effective", "", "", "", "Very effective"), name = "Belief in efficacy of the imposed measures") +
  scale_y_continuous(name="Frequency", limits = c(0,4)) + scale_fill_brewer(palette="Set1") +
  scale_color_brewer(palette="Set1") + theme_classic() + labs(fill = "Covid result", color = "Covid result") +
  theme(axis.ticks.y = element_blank(), axis.text.y=element_blank(), legend.position = "top", plot.margin =
          unit(c(1, 1, 0, 0), "cm"))

freqriskperception <- ggplot(sum_norm, aes(x = bezorgd, fill=Covid_result)) + geom_density(alpha=0.7) + 
  geom_vline(data=mu, aes(xintercept = meanbezorgd, color=Covid_result), linetype="solid", size=1.5) +
  scale_x_continuous(labels = c("Not concerned", "", "", "", "Very concerned"), name = "Risk perception") +
  scale_y_continuous(name="Frequency", limits = c(0,4)) +  scale_fill_brewer(palette="Set1") +
  scale_color_brewer(palette="Set1") + theme_classic() + labs(fill = "Covid result", color = "Covid result") +
  theme(axis.ticks.y = element_blank(), axis.text.y=element_blank(), legend.position = "top", plot.margin =
          unit(c(1, 1, 0, 0), "cm"))

freqriskperception
freqbehavior
freqeffective
freqtrust

ggplot(newData, aes(x=Leeftijd, y = Geslacht)) + geom_count()
```

```{r more plotting exploratory}
covidage<-ggplot(summary, aes(x=Covid_result, y = Leeftijd)) + geom_violin() + theme_classic() + scale_x_discrete(name="Covid result") + scale_y_continuous(name="Age")

covidgender <- ggplot(summary) +
  aes(x = Geslacht, fill = factor(Covid_result)) +
  geom_bar(position = "fill") + scale_x_discrete(name = "Gender", labels = c("Female", "Male")) +
  scale_fill_brewer(palette="Set1")+
  scale_y_continuous(name = "Proportion") + labs(fill = "Covid result") + theme_classic()

agebehavior <- ggplot(meansage, aes(x=Leeftijd_cat, y = gedrag, fill = Covid_result)) + geom_bar(stat='identity', position ='dodge') + scale_fill_brewer(palette = "Set1") + theme_classic() + scale_y_continuous(name = "Behavior", limits = c(0,0.8), breaks=c(0,0.8), labels = c("Not strict", "Very strict")) + labs(fill = "Covid result") + scale_x_discrete(name = "Age categories (in years)") + coord_flip() + facet_grid(Geslacht ~ .)

ageriskperception <- ggplot(meansage, aes(x=Leeftijd_cat, y = bezorgd, fill = Covid_result)) + geom_bar(stat='identity', position ='dodge') + scale_fill_brewer(palette = "Set1") + theme_classic() + scale_x_discrete(name = "Age categories (in years)") + labs(fill = "Covid result") + scale_y_continuous(name = "Risk perception", limits = c(0,0.7), breaks=c(0,0.7), labels = c("Not concerned", "Concerned")) + coord_flip() + facet_grid(Geslacht ~ .)
  
agemeasures <- ggplot(meansage, aes(x=Leeftijd_cat, y = effectiefmaatregelen, fill = Covid_result)) + geom_bar(stat='identity', position ='dodge') + scale_fill_brewer(palette = "Set1") + theme_classic() + scale_y_continuous(name = "Belief in efficacy of the measures", limits = c(0,0.9), breaks=c(0,0.9), labels = c("Not effective", "Very effective")) +  scale_x_discrete(name = "Age categories (in years)") + labs(fill = "Covid result") + coord_flip() + facet_grid(Geslacht ~ .)

agetrust <- ggplot(meansage, aes(x=Leeftijd_cat, y = vertrouwen, fill = Covid_result)) + geom_bar(stat='identity', position ='dodge') + scale_fill_brewer(palette = "Set1") + theme_classic() + scale_y_continuous(name = "Trust in the government and healthcare", limits = c(0,0.9), breaks=c(0,0.9), labels = c("No trust", "A lot of trust")) +  scale_x_discrete(name = "Age categories (in years)") + labs(fill = "Covid result") + coord_flip() + facet_grid(Geslacht ~ .)

ggplot(sum_norm, aes(x=gedragRIVM, y = gedragExtra)) + geom_smooth(color = "orange") + theme_classic() + scale_color_brewer(palette="Oranges") + scale_x_continuous(name = "Behavior RIVM", limits = c(0,1)) + scale_y_continuous(name = "Behavior Extra", limits = c(0,1))

ggplot(sum_norm, aes(x= effectiefmaatregelen, y = bezorgd, color = Covid_result)) + geom_smooth(method='lm') + theme_classic() + scale_color_brewer(palette="Pastel")

ggplot(sum_norm, aes(x= effectiefmaatregelen, y = vertrouwen, color = Covid_result)) + geom_smooth() + theme_classic() + scale_color_brewer(palette="Dark2")

ggplot(sum_norm, aes(x= bezorgd, y = vertrouwen, color = Covid_result)) + geom_smooth() + theme_classic() + facet_grid(Geslacht ~ Covid_result)

ggplot(sum_norm, aes(x= Leeftijd, y = vertrouwen)) + geom_smooth(method='gam') + theme_classic() + scale_color_brewer(palette="Dark2") + facet_grid(Geslacht ~ Covid_result)

ggplot(sum_norm, aes(x= Leeftijd, y = effectiefmaatregelen)) + geom_smooth(method='gam') + theme_classic() + scale_color_brewer(palette="Dark2") + facet_grid(. ~ Covid_result)

ggplot(sum_norm, aes(x= Leeftijd, y = bezorgd)) + geom_smooth(method='gam') + theme_classic() + scale_color_brewer(palette="Dark2") + scale_y_continuous(breaks = c(0,0.5, 1) , labels = c("Strict adherence","", "No strict adherence")) + facet_grid(Covid_result ~ gedrag_b)

ggplot(sum_norm, aes(x= Leeftijd, y = gedrag)) + geom_smooth(method='gam') + theme_classic() + scale_color_brewer(palette="Dark2") + scale_y_continuous(breaks = c(0,0.5, 1) , labels = c("Strict adherence","", "No strict adherence")) + facet_grid(Covid_result ~ effectiefmaatregelen_b)

ggplot(sum_norm, aes(x = Leeftijd, y = gedrag)) + geom_smooth() + theme_classic() + scale_color_brewer(palette="Dark2") + scale_color_brewer(palette="Dark2") + scale_y_continuous(breaks = c(0,5), labels = c("Strict adherence", "No strict adherence")) + facet_grid(Geslacht ~ Covid_result) 

ggplot(newData, aes(x= bezorgd, y = VertrouwenOverheid, color = Covid_result)) + geom_smooth() + theme_classic() +
  scale_color_brewer(palette = "Dark2")

ggplot(newData, aes(x= bezorgd, y = VertrouwenGezondheidssysteem, color = Covid_result)) + geom_jitter() + theme_classic() + scale_color_brewer(palette = "Dark2") +
  scale_x_continuous(name = "Risk perception", limits = c(1,5)) +
  scale_y_continuous(name = "Trust in healthcare", limits = c(0,5))

ggplot(sum_norm, aes(x = gedragRIVM, y = gedragExtra, color=Covid_result)) + geom_jitter(alpha=0.2) + scale_x_continuous(name = "GedragRIVM") + scale_y_continuous(name="GedragExtra") + theme_classic() + theme(plot.margin=unit(c(0.5,1.2,0.5,0.5),"cm"))

ggplot(newData, aes(x = VertrouwenOverheid, y = VertrouwenGezondheidssysteem)) +
  geom_smooth(method='lm') + facet_grid(. ~ Covid_result) + theme_classic()

ggplot(sum_norm, aes(x = gedrag, y = MinutenAandacht)) +
  geom_violin() + facet_grid(. ~ Covid_result)
install.packages("ggstatsplot")

ggplot(sum_norm, aes(x=gedragRIVM, fill = Covid_result)) + geom_histogram(position='dodge')

ggplot(sum_norm, aes(x=gedragExtra, fill = Covid_result)) + geom_histogram(stat='bin',position='identity', alpha=0.5)
```

```{r create tikz files with the plots}
###################################
options(tikzMetricPackages = c("\\usepackage[utf8]{inputenc}","\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}", "\\usepackage{amssymb}"))

tikzDevice::tikz(file = "./plots/freqbehavior.tex", width = 5, height = 5)
freqbehavior
dev.off()

tikzDevice::tikz(file = "./plots/freqeffective.tex", width = 5, height = 5)
freqeffective
dev.off()

tikzDevice::tikz(file = "./plots/freqriskperception.tex", width = 5, height = 5)
freqriskperception
dev.off()

tikzDevice::tikz(file = "./plots/freqtrust.tex", width = 5, height = 5)
freqtrust
dev.off()

tikzDevice::tikz(file = "./plots/covidage.tex", width = 5, height = 5)
covidage
dev.off()

tikzDevice::tikz(file = "./plots/covidgender.tex", width = 5, height = 5)
covidgender
dev.off()

tikzDevice::tikz(file = "./plots/agebehavior.tex", width = 5, height = 5)
agebehavior
dev.off()

tikzDevice::tikz(file = "./plots/agemeasures.tex", width = 5, height = 5)
agemeasures
dev.off()

tikzDevice::tikz(file = "./plots/ageriskperception.tex", width = 5, height = 5)
ageriskperception
dev.off()

tikzDevice::tikz(file = "./plots/agetrust.tex", width = 5, height = 5)
agetrust
dev.off()
```


## Preprocessing for the analysis 
For the function medDML the treatment variable needs to be binary. 
```{r making it binary variables and preprocessing for analysis}
sum_norm <- sum_norm %>% mutate(effectiefmaatregelen_b = ifelse(effectiefmaatregelen>=mean(effectiefmaatregelen), 1, 0))
sum_norm <- sum_norm %>% mutate(bezorgd_b = ifelse(bezorgd>=mean(bezorgd), 1, 0))
sum_norm <- sum_norm %>% mutate(gedragRIVM_b = ifelse(gedragRIVM>=mean(gedragRIVM), 1, 0))
sum_norm <- sum_norm %>% mutate(gedrag_b = ifelse(gedrag>=mean(gedrag), 1, 0))
sum_norm <- sum_norm %>% mutate(vertrouwen_b = ifelse(vertrouwen>=mean(vertrouwen), 1, 0))
sum_norm <- sum_norm %>% mutate(gedragExtra_b = ifelse(gedragExtra>=mean(gedragExtra), 1, 0))
sum_norm <- sum_norm %>% mutate(outcome_b = ifelse(Covid_result=="pos", 1, 0))

#checking again for na values
any(apply(summary, 2, function(x) any(is.na(x))))
asNumeric <- function(x) as.numeric(x)

matrix <- as.matrix(control %>% mutate_if(is.factor, asNumeric) %>% drop_na())
matrix = normalize(matrix)
```

## The mediation analysis 

```{r analysis with own data trust}
d = as.numeric(sum_norm$vertrouwen_b)
m = as.numeric(sum_norm$gedrag)
y = as.numeric(sum_norm$outcome_b)
x = matrix
output=medDML(y=y,d=d,m=m,x=x, multmed = TRUE)
print('IV=TRUST')
round(output$results,3)
print(output$ntrimmed)
```

```{r analysis with own data effectivemeasures}
d = as.numeric(sum_norm$effectiefmaatregelen_b)
m = as.numeric(sum_norm$gedrag)
y = as.numeric(sum_norm$outcome_b)
x = matrix

output=medDML(y=y,d=d,m=m,x=x, trim =0.05)
round(output$results,10)
print(output$ntrimmed)
```

```{r analysis with own data risk perception}
d = as.numeric(sum_norm$bezorgd_b)
m = as.numeric(sum_norm$gedrag)
y = as.numeric(sum_norm$outcome_b)
x = matrix

output=medDML(y=y,d=d,m=m,x=x, multmed = TRUE, trim=0.05)
round(output$results,10)
print(output$ntrimmed)
```

# Baseline analysis
```{r mediation analysis with trust}
d = as.numeric(sum_norm$vertrouwen_b)
m = as.numeric(sum_norm$gedrag)
y = as.numeric(sum_norm$outcome_b)
model.0 <- glm(y ~ d)
summary(model.0)
model.M <- glm(m ~ d)
summary(model.M)
model.Y <- glm(y ~ d + m)
summary(model.Y)
results <- mediate(model.M, model.Y, treat='d', mediator='m',
                   boot=TRUE, sims=1000)
summary(results)
```

```{r mediation anlaysis with effectief maatregelen}
d = as.numeric(sum_norm$effectiefmaatregelen_b)
m = as.numeric(sum_norm$gedrag)
y = as.numeric(sum_norm$outcome_b)
model.0 <- glm(y ~ d)
summary(model.0)
model.M <- glm(m ~ d)
summary(model.M)
model.Y <- glm(y ~ d + m)
summary(model.Y)
results <- mediate(model.M, model.Y, treat='d', mediator='m',
                   boot=TRUE, sims=1000)
plot(results)
summary(results)
```

```{r mediation anlaysis with risk perception}
d = as.numeric(sum_norm$bezorgd_b)
m = as.numeric(sum_norm$gedrag)
y = as.numeric(sum_norm$outcome_b)
model.0 <- lm(y ~ d)
summary(model.0)
model.M <- lm(m ~ d)
summary(model.M)
model.Y <- lm(y ~ d + m)
summary(model.Y)
results <- mediate(model.M, model.Y, treat='d', mediator='m',
                   boot=TRUE, sims=1000)
plot(results)
summary(results)
```